---
title             : "Justifications of Harm in Political Manifestos"
shorttitle        : "Justifications of Harm"

author: 
  - name          : "Morgan Ballesteros"
    affiliation   : " "
    corresponding : yes    # Define only one corresponding author
    email         : "mballesteros@uchicago.edu"

affiliation:
  - id            : " "
    institution   : "University of Chicago"

abstract: |
  
  
keywords          : "manifestos, harm, justification, political"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
library("stringr")
library("tinylabels")
library("here")
r_refs("r-references.bib")

# Source (load) your function definitions from an .R script in the R/ folder.
# This ensures any functions you wrote in scripts are available in this Quarto session.
source(here::here("Thesis", "score_manifestos.R"))
```

```{r import-processed, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}

# Read the processed scoring output produced earlier (relative path inside your repo).
scores <- readr::read_csv(here::here("Thesis", "data", "processed", "manifesto_scores.csv"))

# Quick structural check for you while developing (column names/types, a few values).
# With results: false, this will not print in the PDF, but it *will* run.
dplyr::glimpse(scores)
```


```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

The theory of dyadic morality suggests that judgments of harm give moral valence to
relationships between an intentional agent and vulnerable patient (Schein & Gray, 2018).
That is, dyadic morality predicts that if one were to see someone (i.e., an agent) intentionally harming another person (i.e., a vulnerable patient), one would judge that this action (i.e., the harming of another person) is immoral or morally wrong. Furthermore, social interactions are composed of moral norms as mutually agreed upon rules as to how citizens ought to behave. Although moral norms (e.g., do not murder) are meant to keep people from damaging others, some justify norm violations — and harm — in pursuit of a goal (e.g., political assassinations), whereas others never condone violence to induce change. Understanding what differentiates violent and non-violent agents is therefore useful in protecting against harmful acts. One source that marks such differences is the political manifesto (i.e., “a written statement declaring publicly the intentions, motives, or views of its issuer”; Merriam-Webster, n.d.). The current study targets differences in linguistic indicators of justifications of harm and integrative complexity between violence-condoning and non-violent manifestos.  
Research on linguistic markers of political violence follows the growing number of far-right
manifestos published online prior to terrorist attacks (Capelos et al., 2025; Ebner et al.,
2025). Some of these studies evaluate manifestos through the fusion-plus-threat model of
violent extremism (Whitehouse, 2018), which predicts that individuals who lump their personal
and group identities together will be more likely to self-sacrifice for their group.
Associated linguistic work suggests that words indicating identity fusion, group threat, and
violence-condoning norms — including justifications of violence — are correlated with 
terrorism (Ebner et al., 2024; Ebner et al., 2025). While the existing literature grounds the
importance of justificatory rhetoric in predicting terrorism, harm has not yet been examined 
as central to moral judgment despite its importance in some theories of moral judgment (e.g.,
the theory of dyadic morality). The present study serves as a real-world case on how emotion 
and reason interact to produce the judgments and justifications necessary for enacting harm. 
Connecting the political manifesto and moral judgment literatures will enlighten respective 
research as to (a) how moral reasoning implicates justification in manifestos and (b) how 
justification operates in the moral reasoning process, each of which could have policy 
implications.
#| do  I keep the above sentence "The present study..." in oultining implications here, or move it?
  Integrative Complexity is a linguistic measure of cognitive complexity composed of two primary components: (a) Differentiation and (b) Integration (Suedfeld et al., 1992). Differentiation indicates how many perspectives one accounts for; Integration indicates the extent to which those differentiated perspectives are synthesized. A useful example is an argumentative essay, which, if it is complex, recognizes a variety of possible arguments for or against some claim, and synthesizes those arguments into one ultimate argument, using parts of each subsidiary argument (Ballesteros, 2024). Integrative complexity shows a robust relationship with political violence (Suedfeld, 2010). However, research using the construct as a predictor of political violence has largely focused on political leaders in respect to events like war and international crises (Suedfeld, 2010). To my knowledge, no study to date has directly linked the integrative complexity of political manifestos with justifications of political violence. Considering the measure’s long-standing precedent in predicting political events, integrating integrative complexity into the manifesto literature could help identify the factors that distinguish between merely violent rhetoric and rhetoric that induces violence, thereby affording greater predictability through an aggregation of said factors.
#| again, should I keep the above sentence: "Considering the measure's..." as it provides reason for the work, but seemingly more for a proposal structure than an actual thesis

# Current Study

The current study contributes to research seeking to identify differences between violent and non-violent political agents through an evaluation of political manifestos in two new ways: (a) by focusing a qualitative analysis on rhetoric about harm as designating moral valence by the theory of dyadic morality and (b) by scoring manifestos on integrative complexity. I seek to aid in answering three related questions: (a) do justifications of harm correspond to changes in individual moral norms?; (b) does integrative complexity predict whether a manifesto will result in violence versus non-violence?; and (c) does integrative complexity predict both temporally near and far instances of violence in relation to political manifestos? I predict that (a) changes in moral norms — as exemplified in the agents’ writings — and subsequent justification of increasingly harmful acts will be associated with a higher likelihood of violence, (b) lower levels of integrative complexity in political manifestos will predict higher likelihoods of violence, and (c) a decrease in integrative complexity will be associated with an individual coming to believe that harm is justified in pursuit of their political goals.

# Method

## Manifesto Selection

This study includes manifestos divided into three categories following Ebner et al. (2025); (a) violent self-sacrificial, (b) ideologically extreme, and (c) moderate. It also affords longitudinal analysis. I selected not only one-off manifestos already analyzed in the aforementioned literature (e.g.,  Anders Behring Breivik’s Manifesto of the Norway Attacks, 2011; Marx & Engels Manifesto of the Communist Party, 1848), but also sets of writings of individual authors across time (e.g., selections of Vladimir Lenin’s letters). Materials include ten manifestos for each of the aforementioned categories, and three longitudinal sets of writings for each category. Incorporating the 15 manifestos previously analyzed in the literature brings the study to a categorical count as follows: (a) nine violent self-sacrificial texts, (b) three ideologically extreme texts, and (c) 3 moderate texts. Each of these texts are one-off samples. To fill out the rest of the ten one-off samples for each category, manifestos were selected according to their place on the political spectrum (i.e., left to right political platforms), individual actor-author texts versus platform-representative texts, and length in order to match each category along these factors. The same factors were considered in choosing longitudinal samples, and selection included qualitatively matching samples across categories.
#| Update manifesto selection examples in brackets.
#| Did I appropriately select manifestos according to their place on the political spectrum as indicated above?

## Measures
  *Linguistic Indicators.* Ebner et al. (2025) supplies a list of words that indicate
justifications of violence and, as a higher-order category of linguistic markers, violence
condoning norms. The current study uses the set of words indicating violence condoning
norms with minor alterations. First, indicators like “need for jihaad” (Ebner et al., 2025)
have been replaced with relevant phrasing (e.g., need for war; need for revolution). Second,
specific indicators of talk about harm in dyadic morality (e.g., harm, damage, agent,
patient, intentional) have been added to the list of indicators. 
#| Do I still agree with replacement? If so, I need to update run_scoring again.
	*Automated Integrative Complexity.* Hand-scoring integrative complexity is an involved
process requiring multiple scorers and often hours of work per scored material. In light of
time and manpower limitations, this study instead uses automated integrative complexity
(Conway et al., 2014) to score materials, a measure that uses linguistic markers of
differentiation and integration to construct an automated account of integrative complexity.
While automated integrative complexity does not correlate perfectly with hand scores 
(document-level correlation of .82, paragraph-level correlation of .46, Conway et al., 
2020), it is an increasingly validated measure in its own right and often used as an 
efficient and effective replacement for hand scoring. This study uses the document-level
scores for all analyses.  
#| I forgot to do this part in time for integrative data project; IGNORE AUTOMATED INTEGRATIVE COMPLEXITY PARTS!
	*Violence.* Previously analyzed manifestos are easily delineated as to whether they cause 
violence; some manifestos precede lone-actor attacks by their author, the other manifestos 
did not precede violence enacted by their author. The present study maintains this 
distinction. However, it also includes platform-based violence by identifying institutional 
violence in direct causal relationship with the philosophies put forward by corresponding 
manifestos. For example, a manifesto by a political leader that accompanies a war (e.g., the
Russian Imperial Manifesto of 1914 by Tzar Nicholas II announcing Russia’s entry into World 
War One; Golder, 1927) would be counted as violent; a manifesto espoused by a political 
leader that does not accompany institutional violence by that leader’s institution (e.g., 
John F. Kennedy’s "Ich bin ein Berliner") would be counted as non-violent.
	
## Analyses

Following Ebner et al. (2024), I will measure the prevalence of various categories of 
linguistic markers indicating violence condoning norms (i.e., justification of violence, 
martyrdom narrative, hopelessness of alternatives, calls to violence) and harm as a 
percentage of occurrence relative to total sentence count. These prevalence rates will be 
compared across manifesto categories (i.e., violent versus non-violent, extreme versus 
non-extreme) to replicate Ebner et al. (2024) results suggesting that justification of 
violence is a significant predictor between categories. I will create .zip files for each 
selected manifesto and run them through automated integrative complexity, which will give 
scores for integrative complexity, differentiation, and integration. These scores will be 
compared in the same manner as prevalence rates. I will also run regression analyses for 
longitudinal manifesto sets in order to understand the effect changes in integrative 
complexity have on likelihood of violence. I will repeat these analyses for linguistic 
markers of harm on said longitudinal sets.
#| Need to include other categories of markers of violence condoning norms, perhaps even splitting the markers into categories such that I can examine prevalence of individual markers, marker categories, and violence condoning norms all together.
#| Probably need to rewrite this section, at least by tense, but also distribute in different areas of the methods.
	
# Discussion

In line with my aforementioned hypotheses, I predict that (a) greater prevalence of
linguistic indicators of harm will be associated with higher a likelihood of political 
violence, (b) lower integrative complexity, differentiation, and integration scores will be 
associated with a higher likelihood of political violence, and (c) decreasing integrative 
complexity, differentiation, and integration scores longitudinally will be associated with 
higher likelihoods of political violence. The proposed study adds to the ongoing literature 
on linguistic markers of political violence and on the predictive power of integrative 
complexity for major political events. New to the linguistic marker literature is the 
inclusion of longitudinal data, which is useful in that many of the far-right manifestos 
studied in the literature are posted online just prior to or after acts of violence, meaning
that they are limited in their capacity to serve as practically useful predictors of 
terrorism because those who protect against said violence cannot find and act against such 
terrorism quickly enough. Integrative complexity is robust in its long-term predictive 
power, meaning that an analysis of political platform manifestos can be tracked well before 
violence on their account occurs. Politically oriented research using integrative complexity
has not yet looked at lone actor terrorism to my knowledge, instead focusing on political 
leaders’ textual data for its greater accessibility. However, if evaluating manifestos that 
are published shortly before acts of political violence and manifestos that ground political
platforms predict violence, both in the long- and short-term, then integrative complexity 
can become a powerful tool in tracking and stopping what are increasingly prevalent lone 
actor violent acts. This study thus aims to help public safety policy makers understand the 
precursors of political violence and intervene before tragedy strikes.
#| Can use this for guide when witing actual discussion, so hold of on deleting.

## Limitations

The availability of political manifestos is the primary limitation of the proposed study. 
Relevant prior literature on linguistic markers in political manifestos (i.e., Capelos et 
al., 2025; Ebner et al., 2024; Ebner et al., 2025) share sets of either six or 15 
manifestos, citing issues including “the limited availability of terrorist manifestos, 
access restrictions to such documents, as well as researcher time considerations” (Ebner et 
al., 2024). The current study delimits time constraints by using R to automate prevalence 
rates and automated integrative complexity to measure integrative complexity, but manifesto 
accessibility remains an issue. To gain statistical power, I will not limit data selection 
to terrorist manifestos, but will also include writings that contributed to the development 
of political platforms (i.e., The Manifesto of the Communist Party, which is used in prior 
literature, and Vladimir Lenin’s What Is To Be Done?, a novel addition). Thus, access to 
such works through sites like the Marxists Internet Archive provides a wide range of 
writings to choose from, both as select instances and as pieces of a longitudinal series.
#| Can use this for guide when witing actual Limitations, so hold of on deleting.

##Future Directions


## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::