---
title             : "**Justifications of Dyadic Harm in Political Manifestos**"
shorttitle        : "Justifying Dyadic Harm"

author: 
  - name          : "Morgan Ballesteros"
    affiliation   : " "
    corresponding : yes    # Define only one corresponding author
    email         : "mballesteros@uchicago.edu"

affiliation:
  - id            : " "
    institution   : "University of Chicago"

abstract: |
  
  
keywords          : "manifestos, harm, justification, political"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

header-includes:
  - \usepackage{placeins}
  - \usepackage{flafter}

block-headings    : yes

classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(papaja)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(readr)
  library(stringr)
  library(tinylabels)
  library(here)
  library(readxl)
  library(janitor)
  library(broom)
  library(emmeans)
  library(performance)
  library(lmtest)
  library(sandwich)
  library(nnet)
  library(lme4)
})




# Reproducibility checks for key project files.
required_files <- c(
  "run_scoring.R",
  "score_manifestos.R",
  "r-references.bib",
  "data/markers/marker_dictionary.csv"
)

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    paste(
      "The following required files are missing from the thesis directory:",
      paste(missing_files, collapse = ", ")
    )
  )
}

raw_dirs <- c("data/raw", "data/raw/longitudinal")

if (!all(dir.exists(raw_dirs))) {
  stop(
    paste(
      "The following required raw-data directories are missing:",
      paste(raw_dirs[!dir.exists(raw_dirs)], collapse = ", ")
    )
  )
}
normalize_id <- function(x) {
  x |>
    as.character() |>
    tolower() |>
    stringr::str_replace_all("\\.(pdf|csv|txt|xlsx|xls)$", "") |>
    stringr::str_replace_all("pdfbookone$", "") |>
    stringr::str_replace_all("pdf$", "") |>
    stringr::str_replace_all("[^a-z0-9]", "")
}
```

```{r table-format-helpers, include=FALSE}
fmt_num2 <- function(x) {
  ifelse(is.na(x) | is.nan(x), "", sprintf("%.2f", round(x, 2)))
}

fmt_p_apa <- function(p) {
  dplyr::case_when(
    is.na(p) | is.nan(p) ~ "",
    p < .001 ~ "< .001",
    TRUE ~ sub("^0", "", sprintf("%.3f", round(p, 3)))
  )
}

fmt_ci2 <- function(lo, hi) {
  dplyr::if_else(
    is.na(lo) | is.na(hi) | is.nan(lo) | is.nan(hi),
    "",
    sprintf("[%.2f, %.2f]", round(lo, 2), round(hi, 2))
  )
}
```



```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# Ensure processed scoring files exist
# ------------------------------------------------------------
# Purpose:
#   Guarantee that processed scoring outputs exist before analysis proceeds.
#
# Reproducibility logic:
#   If processed files are absent, run_scoring.R is executed automatically.
#   The scoring pipeline should read the raw manifesto corpus from the
#   project's text-first input structure and regenerate processed outputs.

processed_file <- "data/processed/manifesto_scores_groups.csv"

if (!file.exists(processed_file)) {
  message("Processed scoring files not found. Running scoring pipeline...")
  source("run_scoring.R")
  message("Scoring pipeline complete.")
}

# Confirm that the pipeline produced the expected outputs.
expected_outputs <- c(
  "data/processed/manifesto_scores_markers.csv",
  "data/processed/manifesto_scores_groups.csv",
  "data/processed/manifesto_scores_categories.csv"
)

missing_outputs <- expected_outputs[!file.exists(expected_outputs)]

if (length(missing_outputs) > 0) {
  stop(
    paste(
      "The scoring pipeline did not produce the expected files:",
      paste(missing_outputs, collapse = ", ")
    )
  )
}

# Load processed outputs
marker_scores <- read_csv(
  "data/processed/manifesto_scores_markers.csv",
  show_col_types = FALSE
)

group_scores <- read_csv(
  "data/processed/manifesto_scores_groups.csv",
  show_col_types = FALSE
)

category_scores <- read_csv(
  "data/processed/manifesto_scores_categories.csv",
  show_col_types = FALSE
)

```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

The theory of dyadic morality suggests that judgments of harm give moral valence to
relationships between an intentional agent and vulnerable patient [@schein2018theory].
That is, dyadic morality predicts that if one were to see someone (i.e., an agent) intentionally harming another person (i.e., a vulnerable patient), one would judge that this action (i.e., the harming of another person) is immoral or morally wrong. Furthermore, social interactions are composed of moral norms as mutually agreed upon rules as to how citizens ought to behave. Although moral norms (e.g., do not murder) are meant to keep people from damaging others, some justify norm violations — and harm — in pursuit of a goal (e.g., political assassinations), whereas others never condone violence to induce change. Understanding what differentiates violent and non-violent agents is therefore useful in protecting against harmful acts. One source that marks such differences is the political manifesto (i.e., “a written statement declaring publicly the intentions, motives, or views of its issuer”; Merriam-Webster, n.d.). The current study targets differences in linguistic indicators of justifications of harm and integrative complexity between violence-condoning and non-violent manifestos.  
Research on linguistic markers of political violence follows the growing number of far-right
manifestos published online prior to terrorist attacks [@capelos2025today; @ebner2025there]. Some of these studies evaluate manifestos through the fusion-plus-threat model of
violent extremism [@whitehouse2018dying], which predicts that individuals who lump their personal
and group identities together will be more likely to self-sacrifice for their group.
Associated linguistic work suggests that words indicating identity fusion, group threat, and
violence-condoning norms — including justifications of violence — are correlated with 
terrorism [@ebner2024measuring; @ebner2025there]. While the existing literature grounds the
importance of justificatory rhetoric in predicting terrorism, harm has not yet been examined 
as central to moral judgment despite its importance in some theories of moral judgment (e.g.,
the theory of dyadic morality). The present study serves as a real-world case on how emotion 
and reason interact to produce the judgments and justifications necessary for enacting harm. 
Connecting the political manifesto and moral judgment literatures will enlighten respective 
research as to (a) how moral reasoning implicates justification in manifestos and (b) how 
justification operates in the moral reasoning process, each of which could have policy 
implications.  
Integrative Complexity is a linguistic measure of cognitive complexity composed of two primary components: (a) Differentiation and (b) Integration [@suedfeld199227]. Differentiation indicates how many perspectives one accounts for; Integration indicates the extent to which those differentiated perspectives are synthesized. A useful example is an argumentative essay, which, if it is complex, recognizes a variety of possible arguments for or against some claim, and synthesizes those arguments into one ultimate argument, using parts of each subsidiary argument [@ballesteros2025pcc]. Integrative complexity shows a robust relationship with political violence [@suedfeld2010cognitive]. However, research using the construct as a predictor of political violence has largely focused on political leaders in respect to events like war and international crises [@suedfeld2010cognitive]. To my knowledge, no study to date has directly linked the integrative complexity of political manifestos with justifications of political violence. Considering the measure’s long-standing precedent in predicting political events, integrating integrative complexity into the manifesto literature could help identify the factors that distinguish between merely violent rhetoric and rhetoric that induces violence, thereby affording greater predictability through an aggregation of said factors.

## Current Study

The current study contributes to research seeking to identify differences between violent and non-violent political agents through an evaluation of political manifestos in two new ways: (a) by focusing a qualitative analysis on rhetoric about harm as designating moral valence by the theory of dyadic morality and (b) by scoring manifestos on integrative complexity. I seek to aid in answering three related questions: (a) do justifications of harm correspond to changes in individual moral norms?; (b) does integrative complexity predict whether a manifesto will result in violence versus non-violence?; and (c) does integrative complexity predict both temporally near and far instances of violence in relation to political manifestos? I predict that (a) changes in moral norms — as exemplified in the agents’ writings — and subsequent justification of increasingly harmful acts will be associated with a higher likelihood of violence, (b) lower levels of integrative complexity in political manifestos will predict higher likelihoods of violence, and (c) a decrease in integrative complexity will be associated with an individual coming to believe that harm is justified in pursuit of their political goals.

# Method

## Manifesto Selection

This study includes manifestos divided into three categories following @ebner2025there; (a) violent self-sacrificial, (b) ideologically extreme, and (c) moderate. It also affords longitudinal analysis. I selected not only one-off manifestos already analyzed in the aforementioned literature (e.g.,  Anders Behring Breivik’s Manifesto of the Norway Attacks, 2011; Marx & Engels Manifesto of the Communist Party, 1848), but also sets of writings of individual authors across time (e.g., selections of Vladimir Lenin’s letters). Materials include ten manifestos for each of the aforementioned categories, and three longitudinal sets of writings for each category. Incorporating the 15 manifestos previously analyzed in the literature brings the study to a categorical count as follows: (a) nine violent self-sacrificial texts, (b) three ideologically extreme texts, and (c) 3 moderate texts. Each of these texts are one-off samples. To fill out the rest of the ten one-off samples for each category, manifestos were selected according to their place on the political spectrum (i.e., left to right political platforms), individual actor-author texts versus platform-representative texts, and length in order to match each category along these factors. The same factors were considered in choosing longitudinal samples, and selection included qualitatively matching samples across categories.
<!-- Update manifesto selection examples in brackets.
Did I appropriately select manifestos according to their place on the political spectrum as indicated above?
-->

## Measures

### Linguistic Indicators  

@ebner2025there supplies a list of words that indicate
justifications of violence and, as a higher-order category of linguistic markers, violence
condoning norms. The current study uses the set of words indicating violence condoning
norms with minor alterations. First, indicators like “need for jihaad” [@ebner2025there]
have been replaced with relevant phrasing (e.g., need for war; need for revolution). Second,
specific indicators of talk about harm in dyadic morality (e.g., harm, damage, agent,
patient, intentional) have been added to the list of indicators.  
<!-- Do I still agree with replacement? If so, I need to update run_scoring again.
-->

### Automated Integrative Complexity 

Hand-scoring integrative complexity is an involved
process requiring multiple scorers and often hours of work per scored material. In light of
time and manpower limitations, this study instead uses automated integrative complexity
[@conway2014automated] to score materials, a measure that uses linguistic markers of
differentiation and integration to construct an automated account of integrative complexity.
While automated integrative complexity does not correlate perfectly with hand scores 
[document-level correlation of .82, paragraph-level correlation of .46, @conway2020validating], it is an increasingly validated measure in its own right and often used as an 
efficient and effective replacement for hand scoring. This study uses the document-level
scores for all analyses.  

### Violence 

Previously analyzed manifestos are easily delineated as to whether they cause 
violence; some manifestos precede lone-actor attacks by their author, the other manifestos 
did not precede violence enacted by their author. The present study maintains this 
distinction. However, it also includes platform-based violence by identifying institutional 
violence in direct causal relationship with the philosophies put forward by corresponding 
manifestos. For example, a manifesto by a political leader that accompanies a war (e.g., the
*Russian Imperial Manifesto of 1914* by Tzar Nicholas II announcing Russia’s entry into World 
War One; Golder, 1927) would be counted as violent; a manifesto espoused by a political 
leader that does not accompany institutional violence by that leader’s institution (e.g., 
John F. Kennedy’s *Ich bin ein Berliner*) would be counted as non-violent.

### Analyses

Following @ebner2024measuring, I measure the prevalence of each linguistic marker group as occurrences per 1,000 words (word-count–normalized rates). Marker groups include dyadic harm language (added to reflect dyadic morality), justification for violence, martyrdom narratives, hopelessness of alternatives, role model status, and violent references.

To test whether these linguistic indicators differ across manifesto categories (moderate, ideologically extreme, violent self-sacrificial), I estimate linear models predicting each marker group’s prevalence from category and report category means with confidence intervals.

To directly test the hypothesis that integrative complexity predicts violence versus non-violence, I estimate logistic regression models predicting whether a manifesto is classified as violent (vs. non-violent) from document-level AutoIC integrative complexity scores (and, in expanded models, from the marker-group prevalence measures).

For longitudinal corpora (multiple texts per author over time), I estimate mixed-effects regression models assessing (a) within-author changes in integrative complexity and harm-justification language over time and (b) whether lower integrative complexity is associated with higher harm-justification language within authors. Where violence event dates are available, I additionally examine whether integrative complexity and harm-justification indicators differ in texts written nearer versus farther from the violence event.


# Results 

```{r doc-level-dataset, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Compute document total word count from segment-level word_count
doc_word_counts <- group_scores |>
  dplyr::group_by(doc_id, corpus, author, segment_id) |>
  dplyr::summarise(seg_words = max(word_count, na.rm = TRUE), .groups = "drop") |>
  dplyr::group_by(doc_id, corpus, author) |>
  dplyr::summarise(doc_words = sum(seg_words, na.rm = TRUE), .groups = "drop")

# Aggregate marker-group counts to document level, then compute prevalence per 1,000 words
doc_group_counts <- group_scores |>
  dplyr::group_by(doc_id, corpus, author, marker_group) |>
  dplyr::summarise(count = sum(count, na.rm = TRUE), .groups = "drop") |>
  dplyr::left_join(doc_word_counts, by = c("doc_id", "corpus", "author")) |>
  dplyr::mutate(prevalence = (count / doc_words) * 1000)

# Wide prevalence table for analyses
doc_scores <- doc_group_counts |>
  dplyr::select(doc_id, corpus, author, marker_group, prevalence) |>
  tidyr::pivot_wider(
    names_from = marker_group,
    values_from = prevalence,
    values_fill = 0
  )
```

```{r load-doc-meta, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_meta <- readr::read_csv(
  here::here("data/raw/doc_meta.csv"),
  show_col_types = FALSE
)
```


```{r normalize-docids, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores <- doc_scores |>
  dplyr::mutate(doc_id = normalize_id(doc_id))

doc_meta <- doc_meta |>
  dplyr::mutate(doc_id = normalize_id(doc_id))

```

```{r merge-doc-meta, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores <- doc_scores |>
  dplyr::left_join(
    doc_meta |>
      dplyr::select(doc_id, category),
    by = "doc_id"
  )
```


```{r restrict-to-thesis-docs, echo=FALSE, include=FALSE}
# Preserve the full document-level dataset (needed for longitudinal analyses)
doc_scores_all <- doc_scores

# Thesis “one-off categorized” subset used in cross-sectional category analyses
doc_scores <- doc_scores |>
  dplyr::filter(!is.na(category))
```


```{r check-category-merge, echo=FALSE, include=FALSE}
table(doc_scores$category, useNA = "ifany")
```

```{r inspect-doc-meta-join, echo=FALSE, include=FALSE}
nrow(doc_meta)
nrow(doc_scores)

head(doc_meta$doc_id, 20)
head(doc_scores$doc_id, 20)

table(doc_scores$category, useNA = "ifany")

doc_scores %>%
  dplyr::filter(is.na(category)) %>%
  dplyr::select(doc_id, corpus, author) %>%
  head(20)
```



```{r inspect-doc-meta, echo=FALSE, include=FALSE}
names(doc_meta)
head(doc_meta)
head(doc_scores$doc_id, 20)
head(doc_meta$doc_id, 20)
```

```{r autoic-pipeline, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
# ------------------------------------------------------------
# AutoIC: single, coherent pipeline (import -> normalize -> doc table)
# ------------------------------------------------------------

# 1) ONE stable normalizer: define once, use everywhere.
#    (If you keep your setup() definition, replace it with this version and
#     DO NOT redefine normalize_id later.)


# 2) Discover AutoIC files from all relevant folders
autoic_dirs <- c(
  here::here("data/raw/autoic"),
  here::here("data/raw/longitudinal/autoic")
)
autoic_dirs <- autoic_dirs[dir.exists(autoic_dirs)]

autoic_files <- purrr::map(
  autoic_dirs,
  ~ list.files(
      .x,
      pattern = "\\.(xlsx|xls|csv)$",
      full.names = TRUE,
      recursive = TRUE,
      ignore.case = TRUE
    )
) |>
  unlist(use.names = FALSE)

if (length(autoic_files) == 0) {
  stop("No AutoIC files found in: ", paste(autoic_dirs, collapse = ", "))
}

# 3) Map AutoIC file paths to a corpus label that matches doc_scores$corpus
doc_corpus_levels <- sort(unique(doc_scores$corpus))

# Heuristic: find the label that contains "long"
long_label <- doc_corpus_levels[stringr::str_detect(tolower(doc_corpus_levels), "long")]
nonlong_label <- setdiff(doc_corpus_levels, long_label)

# Fallbacks if the heuristic can't find exactly one of each
if (length(long_label) != 1 || length(nonlong_label) != 1) {
  long_label <- "longitudinal"
  nonlong_label <- "non_longitudinal"
}

infer_corpus_from_path <- function(path) {
  file_low <- tolower(basename(path))

  if (stringr::str_detect(file_low, "longitudinal")) {
    long_label[1]
  } else {
    nonlong_label[1]
  }
}


# 4) Robust reader: supports Excel and CSV.
#    For Excel, prefer the sheet named "Document"; otherwise fall back to sheet 1.
read_autoic_file <- function(path) {
  ext <- tolower(tools::file_ext(path))

  dat <- if (ext %in% c("xlsx", "xls")) {
    sheets <- readxl::excel_sheets(path)
    sheet_to_read <- if ("Document" %in% sheets) "Document" else sheets[1]
    readxl::read_excel(path, sheet = sheet_to_read)
  } else if (ext == "csv") {
    readr::read_csv(path, show_col_types = FALSE)
  } else {
    stop("Unsupported AutoIC file type: ", path)
  }

    dat |>
    janitor::clean_names() |>
    dplyr::mutate(
      source_path = path,
      source_file = basename(path),
      doc_id = normalize_id(document),
      corpus = infer_corpus_from_path(path)
    )
}

# 5) Build the row-level table (one row per file, often one row per doc)
autoic_all <- purrr::map_dfr(autoic_files, read_autoic_file)

# 6) Collapse to a key-unique document-level AutoIC table.
#    This prevents left_join() from multiplying rows if there are duplicates.
mean_or_na <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

autoic_doc <- autoic_all |>
  dplyr::group_by(doc_id, corpus) |>
  dplyr::summarise(
    dplyr::across(dplyr::where(is.numeric), mean_or_na),
    source_file = dplyr::first(source_file),
    .groups = "drop"
  )
autoic_doc |>
  dplyr::select(doc_id, corpus, source_file) |>
  dplyr::distinct() |>
  head(20)


cat("AutoIC rows by corpus:\n")
print(table(autoic_doc$corpus, useNA = "ifany"))

cat("\nExample longitudinal AutoIC rows:\n")
print(
  autoic_doc |>
    dplyr::filter(corpus == "longitudinal") |>
    dplyr::select(doc_id, corpus, source_file, ic) |>
    head(30)
)
```


```{r merge-autoic, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores <- doc_scores |>
  dplyr::left_join(autoic_doc, by = c("doc_id", "corpus"))

doc_scores_all <- doc_scores_all |>
  dplyr::left_join(autoic_doc, by = c("doc_id", "corpus"))

doc_scores_all <- doc_scores_all |>
  dplyr::filter(!(corpus == "longitudinal" & author == "lpm"))
```

```{r check-longitudinal-coverage, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
cat("doc_scores_all rows:", nrow(doc_scores_all), "\n")
cat("Unique corpus values:\n")
print(table(doc_scores_all$corpus, useNA = "ifany"))

cat("\nRows by corpus x has_ic:\n")
print(
  doc_scores_all |>
    dplyr::mutate(has_ic = !is.na(ic)) |>
    dplyr::count(corpus, has_ic)
)

cat("\nExample longitudinal rows:\n")
print(
  doc_scores_all |>
    dplyr::filter(stringr::str_detect(tolower(corpus), "long")) |>
    dplyr::select(doc_id, corpus, author, ic) |>
    head(30)
)
```


```{r diagnose-autoic-join-coverage, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

cat("=== AutoIC join coverage (right after merge-autoic) ===\n")

doc_scores |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_with_ic = sum(!is.na(ic)),
    n_missing_ic = sum(is.na(ic))
  ) |>
  print()

cat("\nBy corpus:\n")
doc_scores |>
  dplyr::group_by(corpus) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_with_ic = sum(!is.na(ic)),
    n_missing_ic = sum(is.na(ic)),
    .groups = "drop"
  ) |>
  print()

cat("\nCorpus levels comparison:\n")
print(sort(unique(doc_scores$corpus)))
print(sort(unique(autoic_doc$corpus)))

cat("\nKey overlap checks:\n")
doc_keys <- doc_scores |> dplyr::distinct(doc_id, corpus)
autoic_keys <- autoic_doc |> dplyr::distinct(doc_id, corpus)

cat("Matches on doc_id only:",
    sum(unique(doc_scores$doc_id) %in% unique(autoic_doc$doc_id)),
    "of", dplyr::n_distinct(doc_scores$doc_id), "\n")

cat("Matches on (doc_id, corpus):",
    nrow(dplyr::inner_join(doc_keys, autoic_keys, by = c("doc_id", "corpus"))),
    "of", nrow(doc_keys), "\n")

cat("\nExample doc_scores keys not found in autoic_doc (first 20):\n")
doc_keys |>
  dplyr::anti_join(autoic_keys, by = c("doc_id", "corpus")) |>
  head(20) |>
  print()

cat("\nExample autoic_doc keys not found in doc_scores (first 20):\n")
autoic_keys |>
  dplyr::anti_join(doc_keys, by = c("doc_id", "corpus")) |>
  head(20) |>
  print()

```


```{r inspect-sheets, echo=FALSE, include=FALSE}
autoic_files[1]
if (tolower(tools::file_ext(autoic_files[1])) %in% c("xlsx", "xls")) {
  readxl::excel_sheets(autoic_files[1])
}
```


```{r inspect-autoic-columns, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
names(autoic_all)
names(autoic_doc)
```


```{r clean-doc-scores, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores <- doc_scores |>
  dplyr::select(
    doc_id, corpus, author,
    dyadic_harm, hopelessness, justification_for_violence,
    martyrdom_narrative, role_model_status, violent_reference,
    category,
    ic, dial, elab,
    ic_differentiation, ic_integration,
    dial_differentiation, dial_integration,
    elab_differentiation, elab_integration
  )
```

```{r prepare-analysis-data, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores <- doc_scores |>
  dplyr::mutate(
    category = factor(category, levels = c("moderate", "extreme", "violent"))
  )

doc_scores_ic <- doc_scores |>
  dplyr::mutate(
    dplyr::across(
      c(dyadic_harm, hopelessness, justification_for_violence,
        martyrdom_narrative, role_model_status, violent_reference),
      \(x) tidyr::replace_na(x, 0)
    )
  ) |>
  dplyr::filter(
    !is.na(ic),
    !is.na(category)
  )

if (nrow(doc_scores_ic) == 0) {
  stop(
    "doc_scores_ic is empty after filtering !is.na(ic) and !is.na(category). ",
    "This almost always means the AutoIC merge produced no matches (ic is all NA). ",
    "Run the diagnose-autoic-join-coverage chunk immediately after merge-autoic."
  )
}
```

```{r inspect-analysis-data, echo=FALSE, include=FALSE}
doc_scores_ic |>
  dplyr::summarise(
    n_rows = dplyr::n(),
    n_missing_ic = sum(is.na(ic)),
    n_missing_category = sum(is.na(category)),
    n_missing_dyadic_harm = sum(is.na(dyadic_harm)),
    n_missing_hopelessness = sum(is.na(hopelessness)),
    n_missing_justification = sum(is.na(justification_for_violence)),
    n_missing_martyrdom = sum(is.na(martyrdom_narrative)),
    n_missing_role_model = sum(is.na(role_model_status)),
    n_missing_violent_reference = sum(is.na(violent_reference))
  )
```





```{r diagnose-model-frame-size-ic, echo=FALSE, include=FALSE}
form_ic <- ic ~ dyadic_harm + hopelessness + justification_for_violence +
  martyrdom_narrative + role_model_status + violent_reference + category

mf <- model.frame(form_ic, data = doc_scores_ic, na.action = na.omit, drop.unused.levels = TRUE)

cat("nrow(doc_scores_ic) =", nrow(doc_scores_ic), "\n")
cat("nrow(model frame used by lm) =", nrow(mf), "\n")
```

```{r check-autoic-integration, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
names(doc_scores)

summary(doc_scores$ic)
sum(is.na(doc_scores$ic))
```



```{r autoic-coverage, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
doc_scores |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_with_ic = sum(!is.na(ic)),
    n_missing_ic = sum(is.na(ic))
  )

doc_scores |>
  dplyr::group_by(corpus) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_with_ic = sum(!is.na(ic)),
    n_missing_ic = sum(is.na(ic)),
    .groups = "drop"
  )
```

```{r diagnose-ic-contrasts, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
form_ic <- ic ~ dyadic_harm + hopelessness + justification_for_violence +
  martyrdom_narrative + role_model_status + violent_reference + category

vars <- all.vars(form_ic)

cat("---- Types in doc_scores_ic ----\n")
print(sapply(doc_scores_ic[, vars, drop = FALSE], class))

cat("\n---- Unique non-NA counts in doc_scores_ic ----\n")
uniq_non_na <- sapply(doc_scores_ic[, vars, drop = FALSE], \(x) dplyr::n_distinct(x, na.rm = TRUE))
print(uniq_non_na)

cat("\n---- Category distribution in doc_scores_ic ----\n")
print(table(doc_scores_ic$category, useNA = "ifany"))

# Mimic what lm() will effectively do: build a model frame and drop NA rows.
# Force category into a factor for inspection, but do not assume it has 2+ observed levels.
tmp <- doc_scores_ic |>
  dplyr::mutate(category = factor(category, levels = c("moderate","extreme","violent")))

mf <- model.frame(form_ic, data = tmp, na.action = na.omit, drop.unused.levels = TRUE)

cat("\n---- Rows used by lm after na.omit ----\n")
cat("nrow(model frame) =", nrow(mf), "\n")

cat("\n---- Category distribution in the model frame ----\n")
print(table(mf$category, useNA = "ifany"))

bad_factors <- names(mf)[sapply(mf, \(x) is.factor(x) && nlevels(x) < 2)]
cat("\n---- Factor predictors with <2 levels in model frame ----\n")
print(bad_factors)

```
```{r verify-types-before-ic-regression, echo=FALSE, include=FALSE}
vars <- c(
  "ic",
  "dyadic_harm", "hopelessness", "justification_for_violence",
  "martyrdom_narrative", "role_model_status", "violent_reference",
  "category"
)

str(doc_scores_ic[, vars, drop = FALSE])

```

```{r category-levels-ic, echo=FALSE, include=FALSE}
print(table(doc_scores_ic$category, useNA = "ifany"))
print(levels(doc_scores_ic$category))
```



```{r regression-ic, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
form_ic <- ic ~ dyadic_harm + hopelessness + justification_for_violence +
  martyrdom_narrative + role_model_status + violent_reference + category

mf <- model.frame(form_ic, data = doc_scores_ic, na.action = na.omit, drop.unused.levels = TRUE)

cat("n used by lm after na.omit =", nrow(mf), "\n")
cat("category distribution in modeling data:\n")
print(table(mf$category, useNA = "ifany"))

if (nrow(mf) == 0) {
  stop(
    "Model frame has 0 rows after na.omit. ",
    "This indicates no complete cases remain. ",
    "Most commonly: ic is missing for all rows due to a failed AutoIC merge. ",
    "Check autoic-coverage and diagnose-autoic-join-coverage."
  )
}

single_level <- names(mf)[sapply(mf, \(x) is.factor(x) && nlevels(x) < 2)]
form_ic2 <- form_ic

if ("category" %in% single_level) {
  warning("`category` has <2 observed levels in the model frame; dropping `category`.")
  form_ic2 <- update(form_ic2, . ~ . - category)
  mf <- model.frame(form_ic2, data = doc_scores_ic, na.action = na.omit, drop.unused.levels = TRUE)
}

model_ic <- lm(form_ic2, data = mf)
summary(model_ic)
```


## Descriptive Statistics 

```{r group-prevalence-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=5.4, fig.height=3.3, out.width = "82%", fig.cap="Mean prevalence of grouped linguistic indicators of harm justification. Note. Bars represent mean grouped marker prevalence per 1,000 words across documents. Error bars represent 95% confidence intervals. Marker groups aggregate lexical indicators defined in the study's marker dictionary."}
# ------------------------------------------------------------
# Plot mean prevalence by marker group with 95% confidence intervals
# ------------------------------------------------------------
# Purpose:
#   Visualize the average prevalence of each marker group across documents,
#   along with uncertainty around each mean.
#
# Why this plot is useful:
#   The raw marker-level output contains too many individual indicators for a
#   readable main-text figure. Aggregating to marker groups makes the figure
#   interpretable while preserving the main substantive distinctions in the data.
#
# Plot logic:
#   First compute the mean prevalence and 95% confidence interval for each
#   marker group, then plot those means as horizontal bars with error bars.
#
# APA formatting choices:
#   - use a single neutral fill rather than group-specific colors
#   - remove the legend because the axis labels already identify each group
#   - use clean axis labels and minimal visual clutter
#   - convert snake_case group labels into publication-readable text

group_plot_data <- group_scores |>
  dplyr::group_by(marker_group) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_prevalence = mean(prevalence, na.rm = TRUE),
    sd_prevalence = sd(prevalence, na.rm = TRUE),
    se_prevalence = sd_prevalence / sqrt(n),
    ci_lower = mean_prevalence - 1.96 * se_prevalence,
    ci_upper = mean_prevalence + 1.96 * se_prevalence,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    marker_group_label = marker_group,
    marker_group_label = stringr::str_replace_all(marker_group_label, "_", " "),
    marker_group_label = stringr::str_to_title(marker_group_label)
  )

ggplot2::ggplot(
  group_plot_data,
  ggplot2::aes(
    x = forcats::fct_reorder(marker_group_label, mean_prevalence),
    y = mean_prevalence
  )
) +
  ggplot2::geom_col(
    width = 0.7,
    fill = "grey70",
    color = "black"
  ) +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = ci_lower, ymax = ci_upper),
    width = 0.2,
    color = "black"
  ) +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = "Marker group",
    y = "Mean prevalence per 1,000 words"
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  ggplot2::theme_classic(base_size = 11) +
  ggplot2::theme(
    legend.position = "none",
    axis.title = ggplot2::element_text(color = "black"),
    axis.text = ggplot2::element_text(color = "black")
  )
```

```{r n-docs-setup, echo=FALSE, message=FALSE, warning=FALSE}
n_docs <- length(unique(group_scores$doc_id))

desc_stats <- group_scores %>%
  group_by(marker_group) %>%
  summarise(
    mean = mean(prevalence, na.rm = TRUE),
    sd = sd(prevalence, na.rm = TRUE),
    min = min(prevalence, na.rm = TRUE),
    max = max(prevalence, na.rm = TRUE),
    .groups = "drop"
  )
```

Descriptive statistics were calculated to examine the prevalence of grouped linguistic indicators of harm justification across the manifesto corpus. Prevalence values represent occurrences of grouped markers per 1,000 words. Across all `r n_docs` manifestos, markers associated with hopelessness were the most common (M = `r round(desc_stats$mean[desc_stats$marker_group == "hopelessness"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "hopelessness"], 2)`), followed by role model status (M = `r round(desc_stats$mean[desc_stats$marker_group == "role_model_status"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "role_model_status"], 2)`), justification for violence (M = `r round(desc_stats$mean[desc_stats$marker_group == "justification_for_violence"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "justification_for_violence"], 2)`), martyrdom narratives (M = `r round(desc_stats$mean[desc_stats$marker_group == "martyrdom_narrative"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "martyrdom_narrative"], 2)`), dyadic harm language (M = `r round(desc_stats$mean[desc_stats$marker_group == "dyadic_harm"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "dyadic_harm"], 2)`), and references to violent actors or events (M = `r round(desc_stats$mean[desc_stats$marker_group == "violent_reference"], 2)`, SD = `r round(desc_stats$sd[desc_stats$marker_group == "violent_reference"], 2)`). Minimum and maximum prevalence values ranged from `r round(min(desc_stats$min), 2)` to `r round(max(desc_stats$max), 2)` depending on marker group.  
```{r manifesto-descriptives, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# ------------------------------------------------------------
# Descriptive statistics for manifesto categories
# ------------------------------------------------------------
# Purpose:
#   Summarize the average prevalence of each marker group across the
#   three manifesto categories used in the study.
#
# Why this chunk exists:
#   Before fitting regression models, it is useful to show the raw
#   descriptive pattern in the data. These means provide an easily
#   interpretable overview of how marker prevalence differs across
#   moderate, ideologically extreme, and violent manifestos.
#
# Output:
#   manifesto_descriptives = summary table of grouped marker prevalence,
#   formatted so that marker groups appear as rows and manifesto
#   categories appear as columns.
#
# Interpretation:
#   Values represent average grouped marker prevalence per 1,000 words.

# Convert tidy long data to wide format (one column per marker)
# This is useful for modeling or visualization but the long
# structure remains the canonical tidy representation.

manifesto_descriptives <- doc_scores |>
  dplyr::group_by(category) |>
  dplyr::summarise(
    dyadic_harm = mean(dyadic_harm, na.rm = TRUE),
    justification_for_violence = mean(justification_for_violence, na.rm = TRUE),
    martyrdom_narrative = mean(martyrdom_narrative, na.rm = TRUE),
    hopelessness = mean(hopelessness, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = -category,
    names_to = "marker_group",
    values_to = "mean_prevalence"
  ) |>
  tidyr::pivot_wider(
    names_from = category,
    values_from = mean_prevalence
  ) |>
  dplyr::mutate(
    marker_group = dplyr::recode(
      marker_group,
      "dyadic_harm" = "Dyadic Harm",
      "justification_for_violence" = "Justification for Violence",
      "martyrdom_narrative" = "Martyrdom Narrative",
      "hopelessness" = "Hopelessness"
    )
  ) |>
  dplyr::rename(`Marker Group` = marker_group) |>
  dplyr::mutate(
    dplyr::across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

papaja::apa_table(
  manifesto_descriptives,
  caption = "Mean prevalence of grouped linguistic indicators by manifesto category.",
  note = "Values represent grouped marker prevalence per 1,000 words."
)
```
```{r category-stats-setup, echo=FALSE, message=FALSE, warning=FALSE}
marker_scores_with_cat <- marker_scores %>%
  left_join(doc_meta, by = "doc_id")

category_stats <- marker_scores_with_cat %>%
  group_by(marker_group, category) %>%
  summarise(
    mean = mean(prevalence, na.rm = TRUE),
    .groups = "drop"
  )
```
Mean marker prevalence also differed across manifesto categories (Table \@ref(tab:manifesto-descriptives)). Violent self-sacrificial manifestos exhibited the highest average prevalence of dyadic harm language (M = `r manifesto_descriptives[["violent"]][manifesto_descriptives[["Marker Group"]] == "Dyadic Harm"]`) and justification for violence (M = `r manifesto_descriptives[["violent"]][manifesto_descriptives[["Marker Group"]] == "Justification for Violence"]`), whereas moderate manifestos displayed very low levels of these indicators (dyadic harm M = `r manifesto_descriptives[["moderate"]][manifesto_descriptives[["Marker Group"]] == "Dyadic Harm"]`; justification for violence M = `r manifesto_descriptives[["moderate"]][manifesto_descriptives[["Marker Group"]] == "Justification for Violence"]`). Martyrdom narratives showed the most pronounced difference across categories, appearing most frequently in violent manifestos (M = `r manifesto_descriptives[["violent"]][manifesto_descriptives[["Marker Group"]] == "Martyrdom Narrative"]`) compared with extreme (M = `r manifesto_descriptives[["extreme"]][manifesto_descriptives[["Marker Group"]] == "Martyrdom Narrative"]`) and moderate texts (M = `r manifesto_descriptives[["moderate"]][manifesto_descriptives[["Marker Group"]] == "Martyrdom Narrative"]`). In contrast, markers associated with hopelessness were most prevalent in ideologically extreme manifestos (M = `r manifesto_descriptives[["extreme"]][manifesto_descriptives[["Marker Group"]] == "Hopelessness"]`), followed by moderate manifestos (M = `r manifesto_descriptives[["moderate"]][manifesto_descriptives[["Marker Group"]] == "Hopelessness"]`) and violent manifestos (M = `r manifesto_descriptives[["violent"]][manifesto_descriptives[["Marker Group"]] == "Hopelessness"]`).

## Regression Analyses 

To examine whether manifesto category predicts the prevalence of harm-related linguistic indicators, a series of linear regression models were estimated. Each model predicted the prevalence of a marker group from manifesto category, with moderate manifestos serving as the reference category.

### Dyadic Harm Language 

The regression model predicting dyadic harm language indicated that manifesto category was marginally associated with variation in dyadic harm prevalence, F(2, 28) = 3.10, *p* = .061, R² = .18. Compared with moderate manifestos, ideologically extreme manifestos contained significantly higher levels of dyadic harm language (*b* = 0.16, SE = 0.08, *t* = 2.16, *p* = .040). Violent self-sacrificial manifestos also showed significantly higher prevalence of dyadic harm markers relative to moderate texts (*b* = 0.18, SE = 0.08, *t* = 2.21, *p* = .035).

### Integrative Complexity Predicting Violence


```{r violent-logistic-models, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Binary outcome: violent vs non-violent (moderate + extreme)
dat <- doc_scores_ic |>
  dplyr::mutate(violent = dplyr::if_else(category == "violent", 1L, 0L)) |>
  dplyr::filter(!is.na(ic))

# Standardize continuous predictors: OR per 1 SD change
dat <- dat |>
  dplyr::mutate(
    ic_z = as.numeric(scale(ic)),
    dyadic_harm_z = as.numeric(scale(dyadic_harm)),
    justification_z = as.numeric(scale(justification_for_violence)),
    martyrdom_z = as.numeric(scale(martyrdom_narrative)),
    hopelessness_z = as.numeric(scale(hopelessness))
  )

# Core hypothesis model (Hypothesis b)
m_ic_only <- stats::glm(violent ~ ic_z, data = dat, family = stats::binomial())

tab_ic <- broom::tidy(m_ic_only, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::filter(term != "(Intercept)") |>
  dplyr::mutate(
    Predictor = "Integrative Complexity (z)",
    OR = round(estimate, 2),
    CI = paste0("[", round(conf.low, 2), ", ", round(conf.high, 2), "]"),
    p = dplyr::case_when(
      p.value < .001 ~ "< .001",
      TRUE ~ sprintf("%.3f", p.value)
    )
  ) |>
  dplyr::select(Predictor, OR, CI, p)

papaja::apa_table(
  tab_ic,
  caption = "Logistic regression predicting violent (vs. non-violent) category from integrative complexity.",
  note = "OR reflects a 1 SD increase in standardized integrative complexity."
)


# Expanded model (Hypotheses a + b jointly, but watch sample size)
m_full <- stats::glm(
  violent ~ ic_z + dyadic_harm_z + justification_z + martyrdom_z + hopelessness_z,
  data = dat,
  family = stats::binomial()
)

# APA-ready table for the expanded model: Odds ratios + 95% CI
tab_or <- broom::tidy(m_full, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::filter(term != "(Intercept)") |>
  dplyr::mutate(
    Predictor = dplyr::recode(
      term,
      "ic_z" = "Integrative Complexity (z)",
      "dyadic_harm_z" = "Dyadic Harm (z)",
      "justification_z" = "Justification for Violence (z)",
      "martyrdom_z" = "Martyrdom Narrative (z)",
      "hopelessness_z" = "Hopelessness (z)"
    ),
    OR = round(estimate, 2),
    CI = paste0("[", round(conf.low, 2), ", ", round(conf.high, 2), "]"),
    p = dplyr::case_when(
      p.value < .001 ~ "< .001",
      TRUE ~ sprintf("%.3f", p.value)
    )
  ) |>
  dplyr::select(Predictor, OR, CI, p)

papaja::apa_table(
  tab_or,
  caption = "Logistic regression predicting violent (vs. non-violent) manifesto category from integrative complexity and linguistic indicators.",
  note = "Odds ratios (OR) reflect 1 SD increases in standardized predictors."
)

```


```{r fig-violent-probability, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, fig.cap="Predicted probability of violent (vs. non-violent) category as a function of integrative complexity (95% CI)."}
      
# Predicted probability curve across IC (holding other predictors at 0 = mean, due to z-scoring)
newdat <- tibble::tibble(
  ic_z = seq(min(dat$ic_z, na.rm = TRUE), max(dat$ic_z, na.rm = TRUE), length.out = 100),
  dyadic_harm_z = 0,
  justification_z = 0,
  martyrdom_z = 0,
  hopelessness_z = 0
)

pred <- stats::predict(m_full, newdata = newdat, type = "link", se.fit = TRUE)

plotdat <- newdat |>
  dplyr::mutate(
    p = stats::plogis(pred$fit),
    lo = stats::plogis(pred$fit - 1.96 * pred$se.fit),
    hi = stats::plogis(pred$fit + 1.96 * pred$se.fit)
  )

ggplot2::ggplot(plotdat, ggplot2::aes(x = ic_z, y = p)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2) +
  ggplot2::geom_line() +
  ggplot2::labs(
    x = "Integrative Complexity (z)",
    y = "Predicted P(Violent)"
  ) +
  ggplot2::theme_classic()
```

```{r fig-violent-oddsratios, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, fig.cap="Odds ratios (95% CI) from the logistic model predicting violent (vs. non-violent) category."}
coef_tab <- broom::tidy(m_full, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::filter(term != "(Intercept)") |>
  dplyr::mutate(term = forcats::fct_rev(term))

ggplot2::ggplot(coef_tab, ggplot2::aes(x = estimate, y = term)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbarh(ggplot2::aes(xmin = conf.low, xmax = conf.high)) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed") +
  ggplot2::labs(x = "Odds Ratio (95% CI)", y = NULL) +
  ggplot2::theme_classic()

```

```{r multinomial-sensitivity, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
dat3 <- doc_scores_ic |>
  dplyr::filter(!is.na(category), !is.na(ic)) |>
  dplyr::mutate(
    category = droplevels(category),
    ic_z = as.numeric(scale(ic)),
    dyadic_harm_z = as.numeric(scale(dyadic_harm)),
    justification_z = as.numeric(scale(justification_for_violence)),
    martyrdom_z = as.numeric(scale(martyrdom_narrative)),
    hopelessness_z = as.numeric(scale(hopelessness))
  )

m_multi <- nnet::multinom(
  category ~ ic_z + dyadic_harm_z + justification_z + martyrdom_z + hopelessness_z,
  data = dat3,
  trace = FALSE
)

summ <- summary(m_multi)
coefs <- as.data.frame(summ$coefficients)
ses <- as.data.frame(summ$standard.errors)

out <- dplyr::bind_rows(lapply(1:nrow(coefs), function(k) {
  cls <- rownames(coefs)[k]
  beta <- as.numeric(coefs[k, ])
  se <- as.numeric(ses[k, ])
  term <- colnames(coefs)
  tibble::tibble(
    Comparison = paste0(cls, " vs baseline"),
    Term = term,
    OR = exp(beta),
    CI = paste0("[", exp(beta - 1.96*se), ", ", exp(beta + 1.96*se), "]")
  )
})) |>
  dplyr::filter(Term != "(Intercept)")

papaja::apa_table(
  out,
  caption = "Multinomial sensitivity model predicting 3-category manifesto classification from integrative complexity and marker prevalence.",
  note = "Estimated via nnet::multinom; ORs are exponentiated coefficients."
)

```


```{r logistic-diagnostics, echo=FALSE, include=FALSE}
# Optional lightweight diagnostics that do not require the `see` package

# Predicted probabilities
p_hat <- stats::fitted(m_full)

# Check for extreme fitted values that may indicate separation
range(p_hat, na.rm = TRUE)

# Basic residual summary
summary(stats::residuals(m_full, type = "deviance"))

# Influence check
summary(stats::cooks.distance(m_full))
```



```{r marker-group-regressions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
marker_groups <- c(
  "dyadic_harm",
  "justification_for_violence",
  "martyrdom_narrative",
  "hopelessness"
)

models <- lapply(marker_groups, function(m) {
  stats::lm(stats::as.formula(paste(m, "~ category")), data = doc_scores)
})

names(models) <- marker_groups

regression_table <- lapply(names(models), function(name) {
  model_summary <- summary(models[[name]])
  coefs <- as.data.frame(model_summary$coefficients)
  coefs$term <- rownames(coefs)
  rownames(coefs) <- NULL

  coefs |>
    dplyr::filter(term != "(Intercept)") |>
    dplyr::transmute(
      Outcome = dplyr::case_when(
        name == "dyadic_harm" ~ "Dyadic Harm",
        name == "justification_for_violence" ~ "Justification",
        name == "martyrdom_narrative" ~ "Martyrdom",
        name == "hopelessness" ~ "Hopelessness",
        TRUE ~ name
      ),
      Contrast = dplyr::case_when(
        term == "categoryextreme" ~ "Extreme vs. Moderate",
        term == "categoryviolent" ~ "Violent vs. Moderate",
        TRUE ~ term
      ),
      B = round(Estimate, 3),
      SE = round(`Std. Error`, 3),
      t = round(`t value`, 2),
      p = dplyr::case_when(
        `Pr(>|t|)` < .001 ~ "< .001",
        TRUE ~ sprintf("%.3f", `Pr(>|t|)`)
      )
    )
}) |>
  dplyr::bind_rows()

papaja::apa_table(
  regression_table,
  caption = "Linear regression coefficients predicting grouped marker prevalence from manifesto category.",
  note = "Moderate manifestos served as the reference category."
)
```

```{r manifesto-regressions, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, results='asis'}
# ------------------------------------------------------------
# Regression analyses predicting autoIC outcomes from
# linguistic indicators and manifesto category
# ------------------------------------------------------------
# Purpose:
#   Estimate whether grouped linguistic indicators and manifesto
#   category predict document-level autoIC outcomes.
#
# Interpretation:
#   Positive coefficients indicate that higher prevalence of the
#   predictor is associated with higher values on the corresponding
#   autoIC outcome, controlling for the other predictors in the model.

# AutoIC outcomes to model
ic_vars <- c(
  "ic",
  "dial",
  "elab",
  "ic_differentiation",
  "ic_integration",
  "dial_differentiation",
  "dial_integration",
  "elab_differentiation",
  "elab_integration"
)

# Predictors to include in each model
predictors <- c(
  "justification_for_violence",
  "dyadic_harm",
  "category"
)

# Fit one regression per autoIC outcome
models_ic <- lapply(ic_vars, function(v) {
  dat <- doc_scores |>
    dplyr::filter(!is.na(.data[[v]]))

  stats::lm(
    stats::as.formula(
      paste(v, "~", paste(predictors, collapse = " + "))
    ),
    data = dat
  )
})

names(models_ic) <- ic_vars

# Build a compact coefficient table
regression_table <- lapply(names(models_ic), function(name) {
  model_summary <- summary(models_ic[[name]])
  coefs <- as.data.frame(model_summary$coefficients)
  coefs$term <- rownames(coefs)
  rownames(coefs) <- NULL

  coefs |>
    dplyr::filter(term != "(Intercept)") |>
    dplyr::transmute(
      Outcome = dplyr::case_when(
        name == "ic" ~ "IC",
        name == "dial" ~ "Dial",
        name == "elab" ~ "Elab",
        name == "ic_differentiation" ~ "IC Differentiation",
        name == "ic_integration" ~ "IC Integration",
        name == "dial_differentiation" ~ "Dial Differentiation",
        name == "dial_integration" ~ "Dial Integration",
        name == "elab_differentiation" ~ "Elab Differentiation",
        name == "elab_integration" ~ "Elab Integration",
        TRUE ~ name
      ),
      Predictor = dplyr::case_when(
        term == "justification_for_violence" ~ "Justification for Violence",
        term == "dyadic_harm" ~ "Dyadic Harm",
        term == "categoryextreme" ~ "Extreme vs. Moderate",
        term == "categoryviolent" ~ "Violent vs. Moderate",
        TRUE ~ term
      ),
      B = round(Estimate, 3),
      SE = round(`Std. Error`, 3),
      t = round(`t value`, 2),
      p = dplyr::case_when(
        `Pr(>|t|)` < .001 ~ "< .001",
        TRUE ~ sprintf("%.3f", `Pr(>|t|)`)
      )
    )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(
    dplyr::across(where(is.character), ~ enc2utf8(as.character(.x)))
  )

papaja::apa_table(
  regression_table,
  caption = "Linear regression coefficients predicting document-level autoIC outcomes from linguistic indicators and manifesto category.",
  note = "Moderate manifestos served as the reference category. Coefficients for linguistic predictors represent associations with document-level autoIC outcomes, controlling for the other predictors in the model."
)
```

### Justifications for Violence 

The regression model predicting justifications for violence was not statistically significant overall, F(2, 28) = 1.88, *p* = .171, R² = .12. Neither ideologically extreme manifestos (*b* = 0.05, SE = 0.16, *t* = 0.33, *p* = .744) nor violent manifestos (*b* = 0.30, SE = 0.17, *t* = 1.78, *p* = .085) differed significantly from moderate manifestos in their prevalence of justification language. However, the coefficient for violent manifestos approached conventional levels of statistical significance, suggesting a potential trend toward greater justificatory rhetoric in manifestos associated with violence. 

### Martyrdom Narratives 

The regression model predicting martyrdom narratives was statistically significant, F(2, 28) = 3.94, *p* = .031, R² = .22. Violent self-sacrificial manifestos exhibited significantly higher levels of martyrdom narrative markers compared with moderate texts (*b* = 0.47, SE = 0.19, *t* = 2.50, *p* = .019). In contrast, ideologically extreme manifestos did not differ significantly from moderate manifestos in martyrdom language (*b* = 0.05, SE = 0.17, *t* = 0.28, *p* = .780). 

### Hopelessness 

The regression model predicting hopelessness markers was not statistically significant overall, F(2, 28) = 2.48, *p* = .102, R² = .15. Neither ideologically extreme manifestos (*b* = 1.75, SE = 1.25, *t* = 1.41, *p* = .171) nor violent manifestos (*b* = -0.93, SE = 1.36, *t* = -0.69, *p* = .497) differed significantly from moderate manifestos in the prevalence of hopelessness language. 

### Summary of Findings 

Across models, manifesto category predicted significant differences in two linguistic domains: dyadic harm language and martyrdom narratives. Both ideologically extreme and violent manifestos contained higher levels of dyadic harm language relative to moderate texts, suggesting that framing political issues in terms of intentional harm and victimization may be characteristic of more radical ideological writing. Additionally, violent self-sacrificial manifestos displayed significantly greater use of martyrdom narratives than moderate texts, indicating that rhetorical appeals to sacrifice and heroism may play an important role in texts associated with political violence.

## Longitudinal Analyses

```{r prepare-longitudinal, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Keep only longitudinal documents with AutoIC scores
long_dat <- doc_scores_all |>
  dplyr::filter(stringr::str_detect(tolower(corpus), "long")) |>
  dplyr::filter(!is.na(author)) |>
  dplyr::mutate(
    dplyr::across(
      c(dyadic_harm, hopelessness, justification_for_violence,
        martyrdom_narrative, role_model_status, violent_reference),
      \(x) tidyr::replace_na(x, 0)
    )
  ) |>
  dplyr::filter(!is.na(ic))

# Extract an approximate date from trailing digits in doc_id
# Handles cases like "harris04201999" (mddyyyy or mmddyyyy) and year-only endings.
extract_doc_date <- function(doc_id) {
  vapply(doc_id, function(x) {
    digits <- stringr::str_extract(x, "\\d+$")

    if (length(digits) == 0 || is.na(digits)) {
      return(as.character(as.Date(NA)))
    }

    year <- suppressWarnings(as.integer(stringr::str_sub(digits, -4, -1)))
    md <- stringr::str_sub(digits, 1, -5)

    if (is.na(year)) {
      return(as.character(as.Date(NA)))
    }

    if (is.na(md) || md == "") {
      return(sprintf("%04d-01-01", year))
    }

    if (nchar(md) == 4) {
      month <- suppressWarnings(as.integer(stringr::str_sub(md, 1, 2)))
      day   <- suppressWarnings(as.integer(stringr::str_sub(md, 3, 4)))
    } else if (nchar(md) == 3) {
      month <- suppressWarnings(as.integer(stringr::str_sub(md, 1, 1)))
      day   <- suppressWarnings(as.integer(stringr::str_sub(md, 2, 3)))
    } else if (nchar(md) == 2) {
      month <- suppressWarnings(as.integer(stringr::str_sub(md, 1, 1)))
      day   <- suppressWarnings(as.integer(stringr::str_sub(md, 2, 2)))
    } else {
      return(sprintf("%04d-01-01", year))
    }

    # Fallback if parsed month/day are invalid
    if (is.na(month) || is.na(day) || month < 1 || month > 12 || day < 1 || day > 31) {
      return(sprintf("%04d-01-01", year))
    }

    sprintf("%04d-%02d-%02d", year, month, day)
  }, FUN.VALUE = character(1)) |>
    as.Date()
}

long_dat <- long_dat |>
  dplyr::mutate(doc_date = extract_doc_date(doc_id)) |>
  dplyr::group_by(author) |>
  dplyr::arrange(doc_date, .by_group = TRUE) |>
  dplyr::mutate(
    time_index = dplyr::row_number(),
    time_c = as.numeric(scale(time_index, center = TRUE, scale = FALSE))
  ) |>
  dplyr::ungroup()


if (nrow(long_dat) == 0) {
  cat(
    "No longitudinal documents with AutoIC scores after merge.\n",
    "This means longitudinal AutoIC rows are not matching on (doc_id, corpus).\n",
    "Run the diagnose-autoic-join-coverage chunk and inspect:\n",
    "  - sort(unique(doc_scores_all$corpus)) vs sort(unique(autoic_doc$corpus))\n",
    "  - anti-join examples of missing keys.\n"
  )
} else {
  cat("Longitudinal analysis rows:", nrow(long_dat), "\n")
}

```

```{r fig-long-ic-trajectories, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Longitudinal trajectories of integrative complexity by author (document order)."}
ggplot2::ggplot(long_dat, ggplot2::aes(x = time_index, y = ic, group = author)) +
  ggplot2::geom_line(alpha = 0.7) +
  ggplot2::geom_point(size = 1) +
  ggplot2::facet_wrap(~ author, scales = "free_x") +
  ggplot2::labs(x = "Document order (within author)", y = "AutoIC Integrative Complexity") +
  ggplot2::theme_classic()
```


```{r longitudinal-models, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Hypothesis c (mechanism): lower IC associated with more harm-justification language within authors
m_just_ic <- lme4::lmer(
  justification_for_violence ~ ic + time_c + (time_c | author),
  data = long_dat,
  REML = FALSE
)

# IC trend over time (is there a decline within authors?)
m_ic_time <- lme4::lmer(
  ic ~ time_c + (time_c | author),
  data = long_dat,
  REML = FALSE
)

# Simple fixed-effect tables (Wald approx)
fixef_table <- function(m, model_label) {
  cf <- summary(m)$coefficients
  out <- tibble::tibble(
    Model = model_label,
    Term = rownames(cf),
    B = round(cf[, "Estimate"], 3),
    SE = round(cf[, "Std. Error"], 3),
    t = round(cf[, "t value"], 2)
  )
  out
}

tab_long <- dplyr::bind_rows(
  fixef_table(m_just_ic, "Justification ~ IC + Time"),
  fixef_table(m_ic_time, "IC ~ Time")
) |>
  dplyr::filter(Term != "(Intercept)")

papaja::apa_table(
  tab_long,
  caption = "Longitudinal mixed-effects models testing within-author change and the association between integrative complexity and justification language.",
  note = "Models include random intercepts and random slopes for time by author (lme4::lmer)."
)
```


```{r harris-near-far, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE, results='asis'}
harris_dat <- long_dat |>
  dplyr::filter(author == "harris") |>
  dplyr::filter(!is.na(doc_date)) |>
  dplyr::mutate(
    event_date = as.Date("1999-04-20"),   # <-- EDIT if you are using a different violence date definition
    days_to_event = as.numeric(event_date - doc_date),
    near_6m = days_to_event >= 0 & days_to_event <= 183
  )

# IC as a function of temporal proximity (within Harris)
m_harris_ic <- stats::lm(ic ~ days_to_event, data = harris_dat)
m_harris_just <- stats::lm(justification_for_violence ~ days_to_event, data = harris_dat)

tab_harris <- tibble::tibble(
  Outcome = c("IC", "Justification for Violence"),
  B_days = c(coef(m_harris_ic)[["days_to_event"]], coef(m_harris_just)[["days_to_event"]]),
  SE_days = c(summary(m_harris_ic)$coefficients["days_to_event","Std. Error"],
             summary(m_harris_just)$coefficients["days_to_event","Std. Error"]),
  p_days = c(summary(m_harris_ic)$coefficients["days_to_event","Pr(>|t|)"],
            summary(m_harris_just)$coefficients["days_to_event","Pr(>|t|)"])
) |>
  dplyr::mutate(
    dplyr::across(c(B_days, SE_days), \(x) round(x, 4)),
    p_days = dplyr::case_when(p_days < .001 ~ "< .001", TRUE ~ sprintf("%.3f", p_days))
  )

papaja::apa_table(
  tab_harris,
  caption = "Within-author temporal proximity models for the violent longitudinal case (Harris).",
  note = "Predictor is days remaining until the violence event date; positive B indicates higher outcome farther from the event."
)
```


# Discussion 

The present study examined whether linguistic indicators associated with moral justifications of harm differ systematically across political manifesto types. Specifically, the analysis tested whether grouped linguistic markers related to dyadic harm framing, justification for violence, martyrdom narratives, and expressions of political hopelessness vary across moderate, ideologically extreme, and violent self-sacrificial manifestos. These analyses represent an initial empirical step toward evaluating the broader hypothesis that linguistic indicators of harm justification correspond to an increased likelihood of political violence.  
The results provide partial support for the proposed theoretical framework. Manifestos categorized as ideologically extreme and violent self-sacrificial exhibited significantly higher levels of dyadic harm language compared with moderate manifestos. In other words, texts associated with more radical ideological positions more frequently framed political conflict in terms of identifiable agents causing harm to identifiable victims.  
This pattern aligns with research in moral psychology emphasizing the centrality of harm perception in moral judgment. In particular, the theory of dyadic morality proposes that moral cognition is structured around a perceived relationship between a moral agent who causes harm and a moral patient who suffers it. According to this framework, moral judgments frequently emerge from narratives that identify perpetrators, victims, and causal responsibility. The elevated prevalence of dyadic harm language in extreme and violent manifestos suggests that such texts more frequently construct political grievances in explicitly moralized terms, portraying social or political conflict as situations in which identifiable actors harm identifiable victims.  
However, the remaining linguistic indicators examined in this study (i.e., justifications for violence, martyrdom narratives, and expressions of political hopelessness) did not differ significantly across manifesto categories. Although these markers were intended to capture rhetorical patterns that legitimize or normalize violence, the results suggest that such language may be present across ideological texts more broadly. Radical political discourse may therefore share similar rhetorical structures regardless of whether it ultimately culminates in violent action.  
One interpretation of this pattern is that linguistic indicators of harm justification may reflect ideological radicalization more generally rather than violence-producing ideologies specifically. The presence of such rhetoric may represent an early stage of moral framing in which political conflicts are interpreted through narratives of victimization and moral grievance. Whether such narratives translate into violent action may depend on additional psychological, cognitive, or situational factors that extend beyond linguistic framing alone.  
Importantly, the present analysis represents only an initial component of the broader research design motivating this project. The theoretical framework underlying the study also proposes that integrative complexity, particularly levels of differentiation and integration in reasoning, may play a key role in determining whether ideological beliefs translate into violent action. Previous research in political psychology has demonstrated that integrative complexity can predict major political events, including diplomatic conflict and war escalation. However, its application to the study of lone-actor political violence remains relatively underexplored.  
Future analyses incorporating automated integrative complexity (AutoIC) scores will allow for direct testing of the remaining hypotheses of this project. Specifically, the study predicts that lower levels of integrative complexity will be associated with a higher likelihood that a manifesto corresponds with political violence. Additionally, longitudinal analyses of texts produced by the same political actors may reveal whether declining levels of integrative complexity precede increasing justification of harm and eventual violent action.

## Limitations 

Several limitations should be considered when interpreting the present findings. The primary limitation concerns the availability of political manifestos. Prior research examining linguistic markers in extremist manifestos has relied on relatively small corpora. For example, studies in this area often analyze sets of six to fifteen manifestos due to the limited availability of such documents, restrictions on access to extremist materials, and the time-intensive nature of manual coding procedures [@capelos2025today; @ebner2024measuring; @ebner2025there]. The present study addresses some of these practical limitations by automating both linguistic prevalence analysis and integrative complexity measurement using computational tools in R. Nevertheless, the accessibility of relevant texts remains a constraint on corpus size.  
To increase statistical power and broaden the scope of analysis, the present study therefore includes not only terrorist manifestos but also ideological writings that contributed to the development of political platforms. For instance, texts such as *The Manifesto of the Communist Party* — which has been included in prior linguistic studies of political manifestos — are analyzed alongside writings such as Vladimir Lenin’s *What Is to Be Done?*, which has not typically been included in this literature. Expanding the dataset to include ideological platform writings allows for the examination of rhetorical patterns across a wider range of political texts while also enabling longitudinal analysis of authors whose writings span multiple time periods.  
However, this strategy also introduces conceptual complexity. Political platform writings may differ from terrorist manifestos in important ways, including their intended audience, rhetorical purpose, and historical context. As a result, the linguistic markers observed in such texts may reflect broader ideological discourse rather than mechanisms uniquely associated with political violence.  
A second limitation concerns the dictionary-based linguistic approach used in the present analysis. Although dictionary methods provide transparency and replicability, they inevitably simplify the semantic richness of political rhetoric. Certain forms of moral justification may therefore occur in ways that are not captured by the specific lexical indicators included in the marker dictionary.  
Finally, the current analysis does not yet incorporate the integrative complexity measures that constitute a central component of the overall research design. As a result, the findings presented here should be interpreted as preliminary. The analyses primarily demonstrate the feasibility of the linguistic scoring pipeline and provide an initial exploration of how harm-related rhetorical patterns vary across manifesto categories.

## Future Directions 

Future research can extend the present work in several important ways. Most immediately, incorporating automated integrative complexity measures will allow for direct testing of whether lower levels of cognitive complexity predict the emergence of political violence. Integrative complexity has demonstrated substantial predictive power in analyses of diplomatic communication and political leadership rhetoric. Applying this approach to the study of political manifestos may therefore provide new insights into the cognitive processes that precede ideological violence.  
Second, the inclusion of longitudinal textual data offers an opportunity to examine whether changes in linguistic framing and cognitive complexity occur prior to violent action. Many extremist manifestos studied in prior literature are released immediately before or after acts of violence, limiting their usefulness as predictive tools. In contrast, ideological writings and political platform texts often precede violent acts by substantial periods of time. If measurable linguistic or cognitive markers within such texts reliably predict later violence, they could provide valuable signals for identifying escalating radicalization processes.  
Third, future work could integrate dictionary-based linguistic analysis with more flexible natural language processing techniques. Methods such as semantic embedding models or topic modeling may allow researchers to identify broader patterns of moral and ideological rhetoric that extend beyond predefined lexical indicators.  
More broadly, integrating linguistic analysis with theoretical frameworks from moral psychology may provide a richer understanding of the cognitive processes underlying political violence. If political grievances are frequently framed through dyadic narratives of harm and victimization, examining how these narratives interact with cognitive structures such as integrative complexity may offer important insight into the psychological pathways through which ideological beliefs develop into violent political action.  
Taken together, the present findings represent an initial empirical test of the first hypothesis of this project: that linguistic indicators of harm justification correspond with a higher likelihood of political violence. The elevated prevalence of dyadic harm language in ideologically extreme and violent manifestos provides preliminary support for the claim that moralized narratives of harm and victimization may characterize increasingly radical political discourse. However, the remaining hypotheses — specifically that lower levels of integrative complexity predict political violence and that declining integrative complexity over time precedes increasing justification of harm — remain to be tested once automated integrative complexity measures are incorporated into the dataset. The present analysis therefore serves as a preliminary stage in a broader research program aimed at integrating linguistic indicators of moral reasoning with cognitive measures of ideological complexity in order to better understand the psychological precursors of political violence.

```{r data analysis, echo=FALSE, include=FALSE}
#Data analysis I used `r cite_r("r-references.bib")` for all analyses. The raw manifesto texts extracted from PDF files are not stored in a tidy format because each document appears as a single character string containing many paragraphs and potential marker occurrences. In tidy data, each variable should occupy a column and each observational unit should occupy a row. For the present analysis, the relevant observational unit is a segment of text within a document. The text-processing pipeline therefore converts each document into paragraph-level segments and then expands these segments so that each row represents a document–segment–marker observation. This structure allows marker counts and prevalence rates to be computed using grouped operations and joined with the marker dictionary metadata.
```




\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

\newpage

\clearpage
# Appendix A

```{r group-descriptives, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# ------------------------------------------------------------
# Descriptive statistics for marker groups
# ------------------------------------------------------------
# Purpose:
#   Summarize the prevalence distribution of each marker group across the corpus.
#
# Why grouped descriptives are used:
#   The marker dictionary classifies individual lexical indicators into broader
#   theoretical groups. These grouped summaries are more interpretable than raw
#   marker-level tables and reduce visual clutter in the main text.
#
# Output:
#   group_descriptives = cleaned summary table with central tendency and spread
#                        for each marker group.
#
# Interpretation:
#   Prevalence values represent grouped marker counts per 1,000 words.

group_descriptives <- group_scores |>
  dplyr::group_by(marker_group) |>
  dplyr::summarise(
    Documents = dplyr::n_distinct(doc_id),
    Mean = mean(prevalence, na.rm = TRUE),
    SD = sd(prevalence, na.rm = TRUE),
    Min = min(prevalence, na.rm = TRUE),
    Max = max(prevalence, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Marker Group` = marker_group,
    `Marker Group` = stringr::str_replace_all(`Marker Group`, "_", " "),
    `Marker Group` = stringr::str_to_title(`Marker Group`)
  ) |>
  dplyr::select(`Marker Group`, Documents, Mean, SD, Min, Max) |>
  dplyr::arrange(dplyr::desc(Mean)) |>
  dplyr::mutate(
    dplyr::across(c(Mean, SD, Min, Max), ~ round(.x, 2))
  )

papaja::apa_table(
  group_descriptives,
  caption = "Descriptive statistics for grouped linguistic indicators of harm justification.",
   note = "Prevalence values represent marker occurrences per 1,000 words. Marker groups aggregate multiple lexical indicators defined in the study's marker dictionary. Document counts indicate the number of manifestos contributing observations to each group.",
  placement = "H"
)
```

\newpage

\clearpage
# Appendix B

```{r manifesto-model-fit, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# ------------------------------------------------------------
# Model fit statistics for manifesto-category regressions
# ------------------------------------------------------------

# Purpose:
#   Provide overall fit information for each regression model in a
#   separate compact table.

model_fit_table <- lapply(names(models), function(name) {
  model_summary <- summary(models[[name]])
  
  tibble::tibble(
    Outcome = dplyr::case_when(
      name == "dyadic_harm" ~ "Dyadic Harm",
      name == "justification_for_violence" ~ "Justification",
      name == "martyrdom_narrative" ~ "Martyrdom",
      name == "hopelessness" ~ "Hopelessness",
      TRUE ~ name
    ),
    R2 = round(model_summary$r.squared, 3),
    Adj_R2 = round(model_summary$adj.r.squared, 3),
    F = round(model_summary$fstatistic[1], 2),
    p = dplyr::case_when(
      stats::pf(
        model_summary$fstatistic[1],
        model_summary$fstatistic[2],
        model_summary$fstatistic[3],
        lower.tail = FALSE
      ) < .001 ~ "< .001",
      TRUE ~ sprintf(
        "%.3f",
        stats::pf(
          model_summary$fstatistic[1],
          model_summary$fstatistic[2],
          model_summary$fstatistic[3],
          lower.tail = FALSE
        )
      )
    )
  )
}) |>
  dplyr::bind_rows() |>

dplyr::rename(
  `R2` = R2,
  `Adj. R2` = Adj_R2
)

papaja::apa_table(
  model_fit_table,
  caption = "Model fit statistics for regressions predicting grouped marker prevalence from manifesto category.",
  note = "R2 = coefficient of determination; Adj. R2 = adjusted coefficient of determination.",
  placement = "H"
)
```
